A Model for Extracting Time Delays from Noisy Time Series Data

### 2013-08-14 doc/Paper1/ms.tex
New title
Working through the intro

\title[Time Delays with Gaussian Processes]
{Inference of Gravitational Lens Time Delays Using Gaussian Processes} 
\author[B. J. Brewer]{Brendon J. Brewer$^1$, Philip J. Marshall$^2$, Gregory Dobler$^3$ \\
$^1$Department of Physics, University of California, Santa Barbara, CA, USA \\
$^2$Department of Physics, University of Oxford, Keble Road, Oxford, OX1 3RH, UK \\
$^3$KITP}

@@

\begin{abstract}

Observations of gravitationally lensed quasars provide a unique method for
measuring cosmological parameters, particularly the Hubble constant $H_0$.
A key part of
this process is the extraction of the time delays from the photometric
monitoring data of the multiple images of the quasar. Quantifying the
uncertainty in this process is challenging, particularly when the time delays
are obscured by the presence of microlensing in the light curves. We present a
fully probabilistic model for the process of inferring the time delays from
the light curve data. In this model, the probability distribution over light
curves is described by a Gaussian Process (GP) whose covariance function
depends on the time delays as well as the properties of the microlensing.
Inferring the parameters of the covariance function from data yields
the time delays, while allowing for a wide range of intrinsic, microlensing, and 
noise-related variability. We test this method extensively on simulated data, 
and find the inferences (including parameter uncertainties) to be robust.
We apply our method to radio and optical monitoring data of the well
studied lenses B1608+656 and RXJ1131$-$123, deriving new time delay estimates complete 
with covariances. The effect on the $H_0$ precision from B1608 is ...

\end{abstract}

@@

However, any inference of $H_0$ from the gravitational lensing method inherits
the uncertainty from the measurements of the time delays themselves. A
thorough investigation of all known sources of uncertainty is an integral part
of measuring cosmological parameters, and thus the purpose of this paper is to
develop and present a reliable method for estimating the uncertainty on the
values of the time delays themselves. The method is Bayesian
\citep{2004kats.book.....O} so the results obtained are probability
distributions for the time delays, and the uncertainties in the time delays
are indicated by the widths of these probability distributions. To describe
the underlying time variation of the quasar and the effects of microlensing on
the light curves, we use Gaussian Processes (GPs), a family of probability
distributions over function space that has computationally convenient
properties \citep{rasmussen}.

@@

Our method is also related to the curve-shifting approach of
\citet{1996A&A...305...97P}, which aims to find those time delays that
minimise the scatter in the shifted light curves. The primary differences
between the GP approach and standard curve shifting are that the GP approach
is fully Bayesian and allows for exploration of the time delay parameter
space, rather than providing a point estimate. Secondly, it is the likelihood
function, rather than the scatter, which guides the inference, making it clear
how uncertainties are to be obtained.

An introduction to Gaussian Processes and their applications can be found in
\citet{rasmussen}. however, in their applications, GPs are usually used as a
prior distribution over the space of functions. Here, GPs describe the
sampling distributions, or the probability of the data given the model
parameters. This is a common situation in astrophysics in fields where the
signals can be modelled by a stochastic process, for example quasar
variability \citep{2009ApJ...698..895K, 2010ApJ...721.1014M, 2011ApJ...735...80Z} and solar-like
oscillations of stars \citep{2009MNRAS.395.2226B}.

@@

\section{The Model for the Inference}

We begin by defining our model for the prior knowledge about how the time
delays give rise rise to the observed data. This results in a probability
distribution for the data given some unknown parameters, i.e. sampling
distributions $p(D|\theta)$. Once specific data is in hand, the sampling
distributions become a likelihood function for the parameters which can be
combined with a prior distribution $p(\theta)$ to give inferences. By Bayes'
rule, the posterior distribution for some parameters $\theta$ given data $D$
is given by:
\begin{equation}
p(\theta|D = D^*) \propto p(\theta)p(D|\theta)|_{D = D^*}
\end{equation}
where $D^*$ is the actual data set that was observed. Throughout this paper,
the data $D$ are a set of measured magnitudes of multiple images of a QSO; see
Figure~\ref{simdata} for an example data set.

The intrinsic quasar variability, in magnitudes, and in whatever band the
observations are carried out in, is described by a function $f(t)$. For a
four-image system, the resulting light curves, if observed without gaps and
noise, would be described by the functions
\begin{eqnarray}
y_1(t) &=& m_1 + f(t) + \mu_1(t) \\
y_2(t) &=& m_2 + f(t - \tau_2) + \mu_2(t) \\
y_3(t) &=& m_3 + f(t - \tau_3) + \mu_3(t) \\
y_4(t) &=& m_4 + f(t - \tau_4) + \mu_4(t) 
\end{eqnarray}
where $\{m_1, m_2, m_3, m_4\}$ are the mean magnitudes of the images,
$\{\tau_2, \tau_3, \tau_4\}$ are the time delays relative to image 1, and the
functions $\{\mu_1(t), \mu_2(t), \mu_3(t), \mu_4(t)\}$ describe the perturbing
effects of microlensing on each of the light curves. The data then consist of
noisy, irregularly sampled measurements of the light curves $\{y_1(t), y_2(t),
y_3(t), y_4(t)\}$.

@@

>>>>

\title[Measuring Time Delays]
{A Model for Extracting Time Delays from Noisy Time Series Data} 
\author[B. J. Brewer]{Brendon J. Brewer$^1$, Philip J. Marshall$^2$\\
$^1$Department of Statistics, The University of Auckland, Private Bag 92019,
Auckland 1142, New Zealand\\
$^2$Department of Physics, University of Oxford, Keble Road, Oxford, OX1 3RH, UK}

@@

\begin{abstract}
Observations of gravitationally lensed quasars provide a unique method for
measuring cosmological parameters, particularly the Hubble constant $H_0$.
A key part of
this process is the measurement of the time delays from the photometric
monitoring of the multiple images of the quasar. The uncertainties in the time
delays propagate to the uncertainties about the cosmological parameters, so it
is important to have an accurate understanding of the size of these uncertainties.
Quantifying the
uncertainty in the time delay measurements is challenging, particularly when the time series
data is affected by microlensing and other complex physical effects. We present a
fully probabilistic model for the process of inferring the time delays from
the light curve data. In this model, the probability distribution over light
curves is described by a Gaussian Process (GP) probability distribution whose covariance function
depends on the time delays as well as the properties of the microlensing.
Inferring the parameters of the covariance function from data yields
the time delays, while allowing for a wide range of intrinsic, microlensing, and 
noise-related variability. We test this method on a simple simulated data set
and a challenging one, 
and find the inferences (including parameter uncertainties) to be robust.
We apply our method to radio and optical monitoring data of the well
studied gravitational lens B1608+656, deriving new time delay estimates and uncertainties. The effect on the $H_0$ precision from B1608 is {\bf no idea yet}.
\end{abstract}

@@

However, any inference of $H_0$ or other cosmological parameters
from the gravitational lensing technique inherits
the uncertainty from the measurements of the time delays themselves. A
thorough investigation of all known sources of uncertainty is an integral part
of measuring cosmological parameters, and thus the purpose of this paper is to
develop and present a reliable method for estimating the uncertainty on the
values of the time delays themselves. The method is Bayesian
\citep{2004kats.book.....O} so the results obtained are probability
distributions over the space of possible solutions for the time delays, and the uncertainties in the time delays
can be summarised by the widths of these probability distributions. To describe
the underlying time variation of the quasar and the effects of microlensing on
the light curves, we use Gaussian Processes (GPs), a family of probability
distributions over function space that has computationally convenient
properties \citep{rasmussen}.

@@

Our method is also related to the curve-shifting approach of
\citet{1996A&A...305...97P}, which aims to find those time delays that
minimise the scatter in the shifted light curves. The primary differences
between the GP approach and standard curve shifting are that the GP approach
is fully Bayesian and allows for exploration of the time delay parameter
space, rather than providing a point estimate of the time delays.
Secondly, it is the likelihood
function, rather than the scatter, which guides the inference, making it clear
how uncertainties are to be obtained. A recent contribution by \citep{2013PhRvD..87l3512H}
described a similar approach to the one presented here. The main differences
are that we implement the model in a Bayesian framework and obtain posterior
samples using MCMC methods, and we also implement several robustness features
(e.g. we allow the possibility of systematically underestimated error bars)
that are not present in the approach of \citet{2013PhRvD..87l3512H}.

An introduction to Gaussian Processes and their applications can be found in
\citet{rasmussen}. In their applications, GPs are usually used as a choice of
prior distribution over the space of functions. In this work, GPs describe the
sampling distributions, or the probability of the data given the model
parameters. This is a common situation in astrophysics in fields where the
signals are complex and the best models we can think of are 
stochastic processes. Two prominent examples are quasar
variability \citep{2009ApJ...698..895K, 2010ApJ...721.1014M, 2011ApJ...735...80Z} and solar-like oscillations of stars \citep{2009MNRAS.395.2226B}.

@@

\section{The Model for the Inference}
We begin by defining our model for the prior knowledge about how the time
delays give rise rise to the observed data. This results in a probability
distribution for the data given some unknown parameters, i.e. the ``sampling
distribution'' (actually distributions, one for each value of the parameters)
$p(D|\theta)$. Once specific data is in hand, the sampling
distributions become a likelihood function for the parameters which can be
combined with a prior distribution $p(\theta)$ to give the posterior distribution.
The posterior distribution for some parameters $\theta$ given data $D$
is given by Bayes' rule:
\begin{equation}
p(\theta|D = D^*) \propto p(\theta)p(D|\theta)|_{D = D^*}
\end{equation}
where $D^*$ is the actual data set that was observed. Throughout this paper,
the data $D$ are a set of measured magnitudes of multiple images of a QSO; see
Figure~\ref{simdata} for an example data set.

The intrinsic quasar variability, in magnitudes, and in whatever band the
observations are carried out in, can be described by a function $q(t)$. For a
two-image system, the resulting light curves, if observed without gaps and
noise, would be described by the functions
\begin{eqnarray}
y_1(t) &=& m_1 + f(t) + \mu_1(t) \\
y_2(t) &=& m_2 + f(t - \tau_2) + \mu_2(t) \\
\end{eqnarray}
where $\{m_1, m_2\}$ are the mean magnitudes of the images,
$\tau_2$ is the time delay of image 2 relative to image 1, and the
functions $\{\mu_1(t), \mu_2(t)\}$ describe any perturbations that are not
shared by both images. A major source of perturbations is the effect of
microlensing on each of the light curves. The data then consist of
noisy, irregularly sampled measurements of the light curves $\{y_1(t), y_2(t)\}$.

The actual data we observe is as follows. We will combine the measurements of all
images into a single vector of $N$ magnitude measurements.
At a set of times
$\{t_1, t_2, ..., t_N\}$, we measure the magnitude of one of the images. These
noisy measurements are denoted $\{Y_1, Y_2, ..., Y_N\}$. We also keep track of
a set of flags $\{k_1, k_2, ..., k_N\}, \forall k_i \in \{1, 2\}$ which specifies
which image was measured. For example, if $k_4 = 2$ then we know that $Y_4$ was
a noisy measurement of image two at time $t_4$.

When we use Bayes' rule, the flags $\{k_1, k_2, ..., k_N\}$ and measurement
times $\{t_1, t_2, ..., t_N\}$ will be subsumed into the prior information. The
data will be the measurements
\begin{eqnarray}
\boldsymbol{D} = \{Y_1, Y_2, ..., Y_N\}.
\end{eqnarray}

The sampling distributions (and hence the likelihood function) will then be
written as a probability distribution for the measurements $\boldsymbol{D}$
given the time delay $\tau$ along with some nuisance parameters.

@@

\bibitem[\protect\citeauthoryear{Hojjati, Kim, 
\& Linder}{2013}]{2013PhRvD..87l3512H} Hojjati A., Kim A.~G., Linder E.~V., 2013, PhRvD, 87, 123512
